---
title: "nlme_rladies"
author: "CCusack"
date: "2023-07-23"
output: 
  html_document:
    toc: true
    toc_depth: 6
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: cosmo
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

- data from Aaron Fisher https://osf.io/nt37e/

# libs dat

```{r message=FALSE, warning=FALSE}
#### my almost always packages ####
if (!require("psych")) {install.packages("psych"); require("psych")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")}

#### data wrangling packages ####
if (!require("purrr")) {install.packages("purrr"); require("purrr")}
if (!require("sjmisc")) {install.packages("sjmisc"); require("sjmisc")}
if (!require("DataCombine")) {install.packages("DataCombine"); require("DataCombine")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("plyr")) {install.packages("plyr"); require("plyr")} 

#### mlm packages ####
if (!require("nlme")) {install.packages("nlme"); require("nlme")} # mixed effects models
if (!require("psychometric")) {install.packages("psychometric"); require("psychometric")} # iccs
if (!require("r2mlm")) {install.packages("r2mlm"); require("r2mlm")} # pseudo-r squared  

dat <- read.csv("mlm_dat.csv")
dim(dat) # 10217 x 32
length(unique(dat$ID)) # 79
```

# data management

## night lags

Here, I:    
1. **add an empty row** after the last beep of the day to account for night lags    
2. **fill in the ID and date** from the row above    
3. **remove the last empty row** for each participant (e.g., remove the empty row btween ID_001 and ID_003)   
4. **renumber beeps**

```{r}
dat <- dat %>% 
  group_split(ID, dayvar) %>% 
  map_dfr(~ add_row(.x, .after = Inf))

dat <- dat %>% tidyr::fill(ID, date)
dat <- dat %>% group_by(ID) %>% slice(-n())

dat <- dat %>% group_by(ID) %>% 
  dplyr::mutate(beepconsec = seq(1:n())) %>% 
  ungroup()
```

## btwn and within

In these code chunks, I create "trait" like variables for the mean. Then, I center them. I create "state" like variables that subtract their trait-like variable from their momentary value. Then, I lag variables (shit down one row.)

### trait vars
```{r}
dat_means <- dat %>% group_by(ID) %>%
summarise_at(vars(energetic:avoid_people), mean, na.rm=TRUE)

colnames(dat_means)[c(2:ncol(dat_means))] <- paste(colnames(dat_means)[c(2:ncol(dat_means))],"trait",sep="_")

dat <- merge(dat, dat_means, by="ID")
```

### center trait vars
```{r}
dat <- dat %>%
  mutate_at(vars(ends_with("trait")),  list(mod = ~ scale(., center=TRUE, scale=FALSE))) %>%
  rename_at(vars(ends_with("_mod")), list(~ paste(gsub("_mod", "_cent", .))))
```

### create state vars
```{r}
# there's definitely a more elegant solution, but here we are
xx = colnames(dat[c(which(colnames(dat)=="energetic"):which(colnames(dat)=="avoid_people"))])
xx <- paste(xx, "state", sep = "_")
dat[xx] <- NA

dat <- dat %>% 
  mutate(energetic_state = energetic - energetic_trait) %>% 
  mutate(enthusiastic_state = enthusiastic - enthusiastic_trait) %>% 
  mutate(content_state = content - content_trait) %>% 
  mutate(irritable_state = irritable - irritable_trait) %>% 
  mutate(restless_state = restless - restless_trait) %>% 
  mutate(worried_state = worried - worried_trait) %>% 
  mutate(guilty_state = guilty - guilty_trait) %>% 
  mutate(afraid_state = afraid - afraid_trait) %>% 
  mutate(anhedonia_state = anhedonia - anhedonia_trait) %>% 
  mutate(angry_state = angry - angry_trait) %>% 
  mutate(hopeless_state = hopeless - hopeless_trait) %>% 
  mutate(down_state = down - down_trait) %>% 
  mutate(positive_state = positive - positive_trait) %>% 
  mutate(fatigue_state = fatigue - fatigue_trait) %>% 
  mutate(tension_state = tension - tension_trait) %>% 
  mutate(concentrate_state = concentrate - concentrate_trait) %>% 
  mutate(accepted_state = accepted - accepted_trait) %>% 
  mutate(threatened_state = threatened - threatened_trait) %>% 
  mutate(ruminate_state = ruminate - ruminate_trait) %>% 
  mutate(avoid_act_state = avoid_act - avoid_act_trait) %>% 
  mutate(reassure_state = reassure - reassure_trait) %>% 
  mutate(procrast_state = procrast - procrast_trait) %>% 
  mutate(hours_state = hours - hours_trait) %>% 
  mutate(difficult_state = difficult - difficult_trait) %>% 
  mutate(unsatisfy_state = unsatisfy - unsatisfy_trait) %>% 
  mutate(avoid_people_state = avoid_people - avoid_people_trait) 
```

### lag
```{r}
namesoriginal <- names(dat[c(which(colnames(dat)=="energetic"):which(colnames(dat)=="avoid_people"))]) # save original names in a vector
nameslag <- paste(names(dat[c(which(colnames(dat)=="energetic"):which(colnames(dat)=="avoid_people"))]), "_lag", sep="") # save new names in a vector

list <- split(dat, dat$ID) # split df by id

list  <- list %>%  # shift all list elements with columns in name original
  map(~shift(.x[namesoriginal],  n = 1, fill=NA, type="lag"))

for(i in seq_along(list)){ # make each list element a dataframe
  list[[i]] <- do.call("cbind", list[[i]])
}

lagdf <- ldply(list, data.frame) # list as dataframe
names(lagdf) <- c("ID", nameslag) # rename vars
lagdf <- lagdf[-1] # remove ID col

dat <- cbind(dat, lagdf) # add lagged vars to dataframe

# clean up environment
rm(list, lagdf, namesoriginal, nameslag, i)

# save df
write.csv(dat, "mlm_dat_processed.csv", row.names = FALSE)
```

# if you start here with preproccesed dat
```{r}
rm(list=ls())
#### my almost always packages ####
if (!require("psych")) {install.packages("psych"); require("psych")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")}

#### data wrangling packages ####
if (!require("purrr")) {install.packages("purrr"); require("purrr")}
if (!require("sjmisc")) {install.packages("sjmisc"); require("sjmisc")}
if (!require("DataCombine")) {install.packages("DataCombine"); require("DataCombine")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("plyr")) {install.packages("plyr"); require("plyr")} 

#### mlm packages ####
if (!require("nlme")) {install.packages("nlme"); require("nlme")} # mixed effects models
if (!require("psychometric")) {install.packages("psychometric"); require("psychometric")} # iccs
if (!require("r2mlm")) {install.packages("r2mlm"); require("r2mlm")} # pseudo-r squared  
dat <- read.csv("mlm_dat_processed.csv")
```


# Random-intercept only
```{r}
#### ICC ####
ICC1.lme(avoid_people_lag, ID, data=dat) # 0.49 variance is at the btwn person level

avoid_RE = lme(avoid_people_lag ~ 1,
            random =~ 1|ID,
            data=dat,
            correlation= corCAR1(form=~beepconsec),
            method="ML", 
            na.action = na.exclude)
summary(avoid_RE)
# random effect (within person diff) = 18.97; between-person diffs = 19.57
intervals(avoid_RE)
VarCorr(avoid_RE)
# intercept = 359.8428
```

# add fixed effects

add in anxiety as both within-person (`anxiety_trait`) and between-person (`anxiety_state`) predictor

Is avoiding people a function of general and momentary worry?

```{r}
avoid_anx1 = lme(avoid_people_lag ~ worried_trait + worried_state +beepconsec,
            random =~ 1|ID,
            data=dat,
            correlation= corCAR1(form=~beepconsec),
            method="ML", # maximum likelihood estimation
            na.action = "na.omit")
summary(avoid_anx1)
# fixed effects: the between person effect is much stronger than within person effect. For each unit increase in momentary worry, avoiding people increases by 0.06. For each unit increase in avg worry, avoiding people increases by 0.55. 
intervals(avoid_anx1)
VarCorr(avoid_anx1)

r2mlm(avoid_anx1, bargraph = TRUE)
# fvm represents the total outcome variance explained by predictors through fixed slopes and random slope variation/covariation and between-person random intercept variation (e.g., total R2 measure; R2total) 0.4903065
# 2) v represents the proportion of outcome variance explained by predictors via random slope variation/covariation (e.g., within-person R2 measure; R2within); 0.00 
# 3) f2 represents the proportion outcome variance explained by between-person variables fixed slopes (e.g., between-person R2 measure; R2between) .25

qqnorm(avoid_anx1, ~ranef(., level=1))
plot(avoid_anx1) # eh not great
```

# add random effect for within-person associations

Let momentary worry vary across persons

```{r}
avoid_anx2 = lme(avoid_people_lag ~ worried_trait + worried_state + beepconsec,
            random =~ 1 + beepconsec + worried_state|ID,
            data=dat,
            correlation= corCAR1(form=~beepconsec),
            method="ML", 
            na.action = "na.omit")

summary(avoid_anx2)
# intervals(avoid_anx2)
# error: "Non-positive definite approximate variance-covariance." too complex. removing beep-consecutive as a random effect

avoid_anx2 = lme(avoid_people_lag ~ worried_trait + worried_state + beepconsec,
            random =~ 1 + worried_state|ID,
            data=dat,
            correlation= corCAR1(form=~beepconsec),
            method="ML", 
            na.action = "na.omit")

summary(avoid_anx2)
# positive slope. pos correlation = .19 correlation btwn ind diff and the within person coupling. People w/ more people avoidance have a stronger association w/ avoidance and momentary worry
intervals(avoid_anx2)
VarCorr(avoid_anx2) 

anova.lme(avoid_anx1, avoid_anx2) # avoid_anx2 has slightly lower AIC and BIC values

r2mlm(avoid_anx2, bargraph = TRUE) 
# fvm = 0.493176037
# v = 0.01
# f2 = 0.25

# although anx2 had lower AIC and BIC values, the more parsimonious model (avoid_anx1) yields similar findings, so I would go with that one rather than 

qqnorm(avoid_anx2, ~ranef(., level=1))
plot(avoid_anx2)
```

# plots

```{r warning=FALSE}
# plotting avoid people by time
dat %>% 
  filter(beepconsec<200) %>% 
  ggplot(aes(beepconsec, avoid_people)) +
  geom_smooth(method='loess', se=FALSE) + 
  # geom_point(color='black',size=.1) +
  labs(x = "Time", y = "Avoid People (0-100)") +
  theme_classic() +
  coord_cartesian(ylim=c(0,100))

# observations w/in person
dat %>%
  filter(beepconsec<200) %>% 
  ggplot(aes(x=beepconsec, y=avoid_people, group=ID, color=ID))+
  geom_smooth(method = "loess", se=FALSE, size=.5)+
  labs(x="Time", y="Avoid People (0-100)") +
  theme_classic() +
  theme(legend.position = "none") +
  coord_cartesian(ylim=c(0,100))

# facet
dat %>%
  filter(beepconsec<200) %>% 
  ggplot(aes(x=beepconsec, y=avoid_people, group=ID, color=ID))+
  geom_smooth(method = "loess", se=FALSE, size=.5)+
  labs(x="Time", y="Avoid People (0-100)") +
  theme_classic() +
  theme(legend.position = "none") +
  coord_cartesian(ylim=c(0,100)) +
  facet_wrap(~ID)  # generally people stay high or stay low
```

